### 2022-03-02 symposium.tex
Background

The Scala standard library contains a package to manipulate regular expressions.
This package is based on Java's implementation, and thus benefits from the high performance of the JVM's regex engine.
The Scala implementation innovates in its presentation: it improves upon Java's solution by providing a more ergonomic API based on pattern matching that is both elegant and concise.
The package's documentation starts with the following example:

@@

In this \thesisOnly{chapter} \paperOnly{paper}, we propose a new design for Scala's regular expression library which provides a type-safe and null-safe mechanism for capturing group extraction.
Our design makes extensive use of match types \paperOnly{\citep{blanvillain2022type}, Scala~3's new feature for type-level programming,} to statically analyze regular expressions during type checking.
We build our interface to mimic Scala's original regular expression API so that Scala programmers can use it as a drop-in replacement and enjoy the additional safety with minimal migration costs.
\thesisOnly{
The source code of our implementation is available online\footnote{\texttt{https://gist.github.com/OlivierBlanvillain/418dcb74cbd2e12eb4a90b7336f40b17}}.
}
\paperOnly{
The source code of our implementation is available in the supplementary material of this submission.
}

\section{Architecture}

@@

\noindent
Unfortunately, this naive solution does not handle regular expressions with nested capturing groups.
For example, in |(A(B)?)|, that solution incorrectly labels the first group as optional.
To overcome this problem, we update our solution to keep track of the number of opening and closing parentheses (|Lvl|), which allows us to differentiate closing parentheses of inner and outer groups%
\footnote{
For simplicity, our presentation treats all non-escaped parentheses as capturing group delimiters.
Our implementation also takes \lstinline!\\Q...\\E! quotations into account and ignores parentheses that appear in character classes.
}:

@@

Nested capturing groups bring another complication to the nullability analysis, which is caused by the interaction between inner and outer groups.
When an outer group is deemed optional, this overrides the nullability of all its inner groups, which also become optional.
For instance, in |(A(B))?|, both the first (|A(B)|) and the second (|B|) capturing group are optional.

@@

In principle, Scala~3's type inference should be able to solve half of that problem with its ability to correlate match types and match expressions\paperOnly{\citep[§ 2.1]{blanvillain2022type}}.
\thesisOnly{
In \Cref{subsec:a-lightweight-form-of-dependent-typing}, we show an example of that ability, where the compiler type checks a small program using a match type.
}
In our case, this mechanism should allow us to get rid of the unsafe cast but does not address the problem of code duplication.
In practice, at the time of writing, this solution does not play well with predefined types from the standard library: the compiler lacks the knowledge necessary to correlate predefined types with their term-level counterpart.
For instance, the compiler does not recognize |CharAt["hello", 0]| as a valid type for the |"hello".charAt(0)|.

@@

\section{Evaluation}

@@

\section{Related Work}

@@

\section{Conclusion}

In this chapter, we introduced a new design for type-safe regular expressions in Scala.
We presented our type-level analysis, which identifies capturing groups and computes their nullability.
We built our library following the design of Scala's original regular expression API, in order to provide a frictionless migration path for programmers interested in the additional type safety.
We evaluated our implementation by running it against the QT3TS test suite and showed that, on those benchmarks, our type-level analysis only has a marginal impact on compilation times.

>>>>

Capturing groups allow programmers to extract substrings matched by parts of a regular expression.
For example, the regex |"A(B)?"| matches both |"A"| and |"AB"|.
In the first case, the capturing group is empty, and in the second case, the capturing group contains |"B"|.

The Scala standard library contains a package to manipulate regular expressions.
This package is based on Java's implementation, and thus benefits from the high performance of the JVM's regex engine.
The Scala implementation innovates in its presentation: it improves upon Java's solution by providing an ergonomic API based on pattern matching that is both elegant and concise.
The package's documentation starts with the following example:

@@

In this \thesisOnly{chapter}\paperOnly{paper}, we propose a new design for Scala's regular expression library which provides a type-safe and null-safe mechanism for capturing group extraction.
Our design makes extensive use of match types\paperOnly{~\citep{blanvillain2022type}, Scala~3's new feature for type-level programming,} to statically analyze regular expressions during type checking.
We build our interface to mimic Scala's original regular expression API so that Scala programmers can use it as a drop-in replacement and enjoy the additional safety with minimal migration costs.

Our presentation is structured as follows:

\begin{itemize}
  \item
\Cref{sec:background}
  \item
\Cref{sec:architecture},~\ref{sec:type-level} and~\ref{sec:term-level}
  \item
\Cref{sec:evaluation}
  \item
\Cref{sec:related}
\Cref{sec:conclusion}

\end{itemize}

\thesisOnly{
The source code of our implementation is available online\footnote{\texttt{https://gist.github.com/OlivierBlanvillain/418dcb74cbd2e12eb4a90b7336f40b17}}.
}
\paperOnly{
The source code of our implementation is available in the supplementary material of this submission.
}

\section{Background}
\label{sec:background}

\subection{Match Types}

In this section, we give a brief introduction to match types by inspecting the following example:

\elemExample
% type Elem[X] = X match
%   case String => Char
%   case Iterable[t] => Elem[t]
%   case Any => X

\noindent
This example defines a type |Elem| parametrized by one type parameter |X|.
The right-hand side is defined in terms of a match on that type parameter---a match type.
A match type reduces to one of its right-hand sides, depending on the type of its scrutinee.
For example, the above type reduces as follows:

\begin{lstlisting}
Elem[String] ~> Char
Elem[Int] ~> Int
Elem[Iterable[Int]] ~> Int
\end{lstlisting}

\noindent
To reduce a match type, the scrutinee is compared to each pattern, one after the other, using subtyping.
For example, although |String| is a subtype of both |String| and |Any|, the |Elem[String]| type reduces to |Char| because the corresponding case appears first.

\subection{The Generic View of Tuples}

% https://www.scala-lang.org/2021/02/26/tuples-bring-generic-programming-to-scala-3.html

\section{Architecture}
\label{sec:architecture}

@@

\noindent
Unfortunately, this naive solution does not handle regular expressions with nested capturing groups.
For example, in |"(A(B)?)"|, that solution incorrectly labels the first group as optional.
To overcome this problem, we update our solution to keep track of the number of opening and closing parentheses (|Lvl|), which allows us to differentiate closing parentheses of inner and outer groups%
\footnote{
For simplicity, our presentation treats all non-escaped parentheses as capturing group delimiters.
Our implementation also takes \lstinline!\\Q...\\E! quotations into account and ignores parentheses that appear in character classes.
}:

@@

Nested capturing groups bring another complication to the nullability analysis, which is caused by the interaction between inner and outer groups.
When an outer group is deemed optional, this overrides the nullability of all its inner groups, which also become optional.
For instance, in |"(A(B))?"|, both the first (|A(B)|) and the second (|B|) capturing group are optional.

@@

In principle, Scala~3's type inference should be able to solve half of that problem with its ability to correlate match types and match expressions\paperOnly{~\citep[§ 2.1]{blanvillain2022type}}.%
\thesisOnly{
In \Cref{subsec:a-lightweight-form-of-dependent-typing}, we show an example of that ability, where the compiler type checks a small program using a match type.
}%
In our case, this mechanism should allow us to get rid of the unsafe cast but does not address the problem of code duplication.
In practice, at the time of writing, this solution does not play well with predefined types from the standard library: the compiler lacks the knowledge necessary to correlate predefined types with their term-level counterpart.
For instance, the compiler does not recognize |CharAt["hello", 0]| as a valid type for the |"hello".charAt(0)|.

@@

\section{Evaluation}
\label{sec:evaluation}

@@

\section{Related Work}
\label{sec:related}

@@

\section{Conclusion}
\label{sec:conclusion}

In this \thesisOnly{chapter}\paperOnly{paper}, we introduced a new design for type-safe regular expressions in Scala.
We presented our type-level analysis, which identifies capturing groups and computes their nullability.
We built our library following the design of Scala's original regular expression API, in order to provide a frictionless migration path for programmers interested in the additional type safety.
We evaluated our implementation by running it against the QT3TS test suite and showed that, on those benchmarks, our type-level analysis only has a marginal impact on compilation times.
